---
title: "01_population_rank_and_vector_operations"
output: html_document
---

# Population Rank and Vector Operations

This notebook demonstrates basic vector operations, ranking, and data manipulation using the murders dataset from the dslabs package.

```{r load-packages, include=FALSE}
# Load required packages for data manipulation and visualization
library(dslabs)
library(tidyverse)
```

## Working with Population Data

### Loading and Exploring the Murders Dataset

```{r load-data}
# Load the murders dataset from dslabs package
data("murders")

# Extract population data and sort it
pop <- murders$population
pop <- sort(pop)

# Display the smallest population value
pop[1]
```

### Finding Minimum Values

```{r find-minimum}
# Alternative way to find the minimum population
min(pop)  # Same result as pop[1]
```

### Finding Index Positions

```{r find-indices}
# Get population data (unsorted)
pop <- murders$population

# Find indices that would sort the population in ascending order
smallest_idx <- order(pop)
smallest_idx[1]  # Index of the state with smallest population

# Get the actual smallest population value using the index
pop[smallest_idx[1]]
```

### Direct Methods for Finding Minimum

```{r direct-minimum}
# Find the index of the minimum population directly
which.min(murders$population)

# Find the state with the minimum population
murders$state[which.min(murders$population)]
```

## Ranking Data

### Creating Population Rankings

```{r population-rankings}
# Create ranks for all states based on population
ranks <- rank(murders$population)

# Create a data frame with state names and their population ranks
my_df <- data.frame(state = murders$state, rank = ranks)

# Display the first few rows
head(my_df)
```

## Working with Named Vectors

### Temperature Conversion Example

```{r temperature-conversion}
# Create temperature data in Celsius for different cities
temp <- c(35, 88, 42, 84, 81, 30)
city <- c("Beijing", "Lagos", "Paris", "Rio de Janeiro", "San Juan", "Toronto")

# Assign city names to temperature values
names(temp) <- city

# Convert Celsius to Fahrenheit using the conversion formula
temp_F <- 5/9 * (temp - 32)
temp_F
```

## Handling Missing Values (NA)

### Working with NA Values

```{r handle-na}
# Load example data with missing values
data("na_example")

# Examine the structure of the data
str(na_example)
```

### Calculating Mean with NA Values

```{r mean-with-na}
# This will return NA because there are missing values
mean(na_example)  # Returns NA
```

### Identifying Missing Values

```{r identify-na}
# Create a logical vector indicating which values are NA
ind <- is.na(na_example)

# Count the number of missing and non-missing values
table(ind)

# Alternative ways to count NA values:
sum(ind)                    # Number of NA values
length(na_example[ind])     # Same as above
```

### Computing Statistics While Excluding NA

```{r exclude-na}
# Method 1: Use na.rm parameter to exclude NA values
mean(na_example, na.rm = TRUE)

# Method 2: Manually exclude NA values using logical indexing
mean(na_example[!ind])
```

## Summary

This notebook covered:

1. **Vector operations**: sorting, finding minimum values, and indexing
2. **Ranking data**: using the `rank()` function to create rankings
3. **Named vectors**: assigning names to vector elements
4. **Missing value handling**: identifying and working with NA values in calculations

These are fundamental operations when working with data in R, especially for data cleaning and exploratory data analysis.

