---
title: "02_Data_Frame_Operations_and_dplyr_Basics"
output: html_document
date: "`r Sys.Date()`"
---

# Data Frame Operations and dplyr Basics

This notebook demonstrates basic data frame operations, filtering, sorting, and summarizing data using the dplyr package with the murders dataset.

```{r load-packages, include=FALSE}
# Load required packages for data manipulation and visualization
library(dslabs)
library(tidyverse)
library(dplyr)
```

## Working with Data Frames

### Loading and Exploring the Dataset

```{r load-explore-data}
# Load the murders dataset
data("murders")

# Examine the structure of the data
str(murders)

# Display the first few rows
head(murders)

# Check dimensions
dim(murders)
```

### Basic Data Frame Operations

```{r basic-operations}
# Access specific columns
murders$state
murders$population

# Use column names to access data
murders[["state"]]

# Display column names
names(murders)
```

## Data Filtering and Selection

### Using Base R for Filtering

```{r base-r-filtering}
# Find states with population less than 1 million
small_states <- murders$state[murders$population < 1000000]
small_states

# Find states in the South region
south_states <- murders$state[murders$region == "South"]
head(south_states)
```

### Using dplyr for Data Manipulation

```{r dplyr-filtering}
# Filter states with population less than 1 million
murders %>% 
  filter(population < 1000000) %>% 
  select(state, population)

# Filter states in the South region
murders %>% 
  filter(region == "South") %>% 
  select(state, region, population)
```

### Multiple Conditions

```{r multiple-conditions}
# Filter states in the South with population > 5 million
murders %>% 
  filter(region == "South" & population > 5000000) %>% 
  select(state, region, population, total)

# Filter states with either very high or very low murder rates
murders %>% 
  mutate(rate = total / population * 100000) %>% 
  filter(rate > 10 | rate < 1) %>% 
  select(state, region, rate) %>% 
  arrange(rate)
```

## Data Sorting and Ranking

### Sorting Data

```{r sorting-data}
# Sort by population (ascending)
murders %>% 
  arrange(population) %>% 
  select(state, population) %>% 
  head()

# Sort by population (descending)
murders %>% 
  arrange(desc(population)) %>% 
  select(state, population) %>% 
  head()
```

### Creating New Variables

```{r new-variables}
# Add murder rate per 100,000 population
murders <- murders %>% 
  mutate(rate = total / population * 100000)

# Display states with highest murder rates
murders %>% 
  arrange(desc(rate)) %>% 
  select(state, region, rate) %>% 
  head(10)
```

## Data Summarization

### Summary Statistics by Region

```{r region-summaries}
# Calculate summary statistics by region
murders %>% 
  group_by(region) %>% 
  summarise(
    total_population = sum(population),
    total_murders = sum(total),
    avg_rate = mean(rate),
    median_rate = median(rate),
    states_count = n()
  )
```

### Finding Extremes

```{r extremes}
# State with highest murder rate
murders %>% 
  filter(rate == max(rate)) %>% 
  select(state, region, rate)

# State with lowest murder rate
murders %>% 
  filter(rate == min(rate)) %>% 
  select(state, region, rate)

# Top 5 safest states (lowest murder rates)
murders %>% 
  arrange(rate) %>% 
  select(state, region, rate) %>% 
  head(5)
```

## Advanced Operations

### Conditional Logic

```{r conditional-logic}
# Categorize states based on murder rate
murders <- murders %>% 
  mutate(
    safety_level = case_when(
      rate < 2 ~ "Very Safe",
      rate < 4 ~ "Safe", 
      rate < 6 ~ "Moderate",
      TRUE ~ "High Risk"
    )
  )

# Count states in each safety category
murders %>% 
  count(safety_level, sort = TRUE)
```

### Working with Text Data

```{r text-operations}
# Find states with "New" in the name
murders %>% 
  filter(str_detect(state, "New")) %>% 
  select(state, region, rate)

# States with names longer than 10 characters
murders %>% 
  filter(str_length(state) > 10) %>% 
  select(state, region) %>% 
  arrange(state)
```

### Data Validation

```{r data-validation}
# Check for missing values
murders %>% 
  summarise_all(~sum(is.na(.)))

# Check data ranges
murders %>% 
  summarise(
    min_pop = min(population),
    max_pop = max(population),
    min_total = min(total),
    max_total = max(total),
    min_rate = min(rate),
    max_rate = max(rate)
  )
```

## Regional Analysis

### Comparing Regions

```{r regional-comparison}
# Create a comprehensive regional summary
regional_analysis <- murders %>% 
  group_by(region) %>% 
  summarise(
    states_count = n(),
    total_population = sum(population),
    total_murders = sum(total),
    avg_rate = mean(rate),
    median_rate = median(rate),
    min_rate = min(rate),
    max_rate = max(rate),
    .groups = 'drop'
  ) %>% 
  arrange(desc(avg_rate))

regional_analysis
```

### Top States by Region

```{r top-states-by-region}
# Find the state with highest murder rate in each region
murders %>% 
  group_by(region) %>% 
  filter(rate == max(rate)) %>% 
  select(region, state, rate) %>% 
  arrange(desc(rate))
```

## Summary

This notebook covered essential data frame operations including:

1. **Basic data exploration**: structure, dimensions, and column access
2. **Filtering data**: using conditions to subset data
3. **Sorting and arranging**: ordering data by different criteria  
4. **Creating new variables**: calculating derived metrics like murder rates
5. **Data summarization**: computing statistics by groups
6. **Advanced operations**: conditional logic, text manipulation, and data validation
7. **Regional analysis**: comparing different geographical regions

These operations form the foundation of data analysis and are essential for exploratory data analysis (EDA) and data preprocessing.
