---
title: "03_tidy_data_and_combing_tables"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  cache = TRUE
)
library(tidyverse)
library(dslabs)
library(knitr)
library(kableExtra)
library(scales)
```

# Introduction

This report provides a complete analysis workflow using the murders dataset from the dslabs package.

## Data Loading and Preparation

```{r}
data("murders")

# Calculate murder rate per 100,000
murders <- murders %>%
  mutate(
    rate = total / population * 100000,
    rate_rounded = round(rate, 1),
    safety_level = case_when(
      rate < 2 ~ "Very Safe",
      rate < 4 ~ "Safe",
      rate < 6 ~ "Moderate",
      TRUE ~ "High Risk"
    )
  )
```

## Exploratory Analysis

### Basic Statistics

```{r}
murders %>%
  select(population, total, rate) %>%
  summary() %>%
  kable(caption = "Summary Statistics") %>%
  kable_styling(bootstrap_options = "striped")
```

### Regional Analysis

```{r}
regional_stats <- murders %>%
  group_by(region) %>%
  summarise(
    avg_rate = mean(rate),
    median_rate = median(rate),
    min_rate = min(rate),
    max_rate = max(rate),
    total_murders = sum(total),
    .groups = "drop"
  )

regional_stats %>%
  kable(caption = "Regional Statistics") %>%
  kable_styling(bootstrap_options = "striped")
```

## Visualization

### Murder Rate Distribution

```{r}
ggplot(murders, aes(x = rate)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Murder Rates",
       x = "Murders per 100,000",
       y = "Number of States") +
  theme_minimal()
```

### Regional Comparison

```{r}
ggplot(murders, aes(x = region, y = rate, fill = region)) +
  geom_boxplot() +
  labs(title = "Murder Rates by Region",
       x = "Region",
       y = "Murders per 100,000") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Top and Bottom States

```{r}
# Generate top_states and bottom_states
top_states <- murders %>%
  arrange(desc(rate)) %>%
  select(state, region, rate) %>%
  head(5)

bottom_states <- murders %>%
  arrange(rate) %>%
  select(state, region, rate) %>%
  head(5)

# Combine top_states and bottom_states into a single data frame with an additional column
extreme_states <- bind_rows(
  top_states %>% mutate(Category = "Highest Murder Rates"),
  bottom_states %>% mutate(Category = "Lowest Murder Rates")
)

# Create a table using kable
extreme_states %>%
  kable(caption = "Extreme States by Murder Rate") %>%
  kable_styling(bootstrap_options = "striped")
```

### Population vs Murder Rate

```{r}
ggplot(murders, aes(x = population/1000000, y = rate)) +
  geom_point(aes(color = region), size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(labels = comma) +
  labs(title = "Population vs Murder Rate",
       x = "Population (millions)",
       y = "Murders per 100,000",
       color = "Region") +
  theme_minimal()
```

### Safety Level Analysis

```{r}
safety_summary <- murders %>%
  count(safety_level) %>%
  mutate(percentage = n / sum(n) * 100)

ggplot(safety_summary, aes(x = safety_level, y = n, fill = safety_level)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), vjust = -0.5) +
  labs(title = "States by Safety Level",
       x = "Safety Level",
       y = "Number of States") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Conclusion

This analysis reveals patterns in murder rates across U.S. states, with notable regional differences. The South has the highest average murder rates while the Northeast has the lowest.

```{r}
sessionInfo()
```

