---
title: "04_low_dim_visualization"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  cache = TRUE
)
library(ggplot2)
library(data.table)
library(magrittr)
library(tidyr)
library(ggrepel)
library(gapminder)
```

# Introduction

This document contains solutions for Exercise Session 4 on low dimensional visualization techniques using ggplot2.

## Tutorial Exercises

### Section 01: Plot Types Matching

- **Boxplot:** Shows distribution and quantiles
- **Bar chart:** Highlights individual values
- **Line chart:** Shows changes over time
- **Scatterplot:** Shows relationships between continuous variables

### Section 02: Reproducing Plots

```{r}
data(mpg)
mpg_dt <- as.data.table(mpg)

# Create requested plot
ggplot(mpg_dt, aes(cty, hwy, color = factor(year))) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = 'lm', se = FALSE) +
  labs(x = "City MPG", y = "Highway MPG", 
       color = "Year",
       title = "Fuel Efficiency Comparison") +
  theme_minimal()
```

### Section 03: Visualizing Distributions (Iris)

#### 1. Load and Examine Data

```{r}
iris_dt <- as.data.table(iris)
head(iris_dt)
tail(iris_dt)
```

#### 2. Histogram of Measurements

```{r}
iris_melt <- melt(iris_dt, id.vars = "Species")

ggplot(iris_melt, aes(value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Distribution of Iris Measurements",
       x = "Measurement Value", y = "Count")
```

#### 3. Varying Bin Numbers

```{r}
# Few bins (oversimplified)
ggplot(iris_melt, aes(value)) +
  geom_histogram(bins = 5, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free")

# Many bins (overly detailed)
ggplot(iris_melt, aes(value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free")
```

#### 4. Boxplot Visualization

```{r}
ggplot(iris_melt, aes(variable, value)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "Measurement Type", y = "Value",
       title = "Boxplot of Iris Measurements")
```

#### 5. Boxplot with Points

```{r}
ggplot(iris_melt, aes(variable, value)) +
  geom_boxplot(outlier.shape = NA) +  # Hide default outliers
  geom_jitter(width = 0.2, alpha = 0.5, size = 1.5) +
  labs(title = "Boxplot with Individual Points")
```

#### 6. Violin-Boxplot Hybrid

```{r}
ggplot(iris_melt, aes(variable, value)) +
  geom_violin(fill = "lightgreen", alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white") +
  labs(title = "Violin Plot with Boxplot Overlay")
```

#### 7. Species-Specific Patterns

```{r}
ggplot(iris_melt, aes(variable, value, color = Species)) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  labs(title = "Measurements by Species") +
  theme(legend.position = "bottom")
```

### Section 04: Continuous Variable Relationships

#### 1. Petal Length vs Width

```{r}
ggplot(iris_dt, aes(Petal.Length, Petal.Width)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Petal Length vs Width",
       x = "Petal Length", y = "Petal Width")
```

#### 2. Species-Specific Correlations

```{r}
# Colored by species
ggplot(iris_dt, aes(Petal.Length, Petal.Width, color = Species)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Petal Dimensions by Species")

# Faceted by species
ggplot(iris_dt, aes(Petal.Length, Petal.Width)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Species, scales = "free") +
  labs(title = "Petal Dimensions by Species (Faceted)")
```

## Homework Exercises

### Section 05: Axes Scaling and Labeling

#### 1. Load and Plot Medals Data

```{r}
medals_dt <- fread('../data/medals.csv')
ggplot(medals_dt, aes(population, total)) +
  geom_point(size = 3) +
  labs(title = "Medals vs Population (Linear Scale)",
       x = "Population", y = "Total Medals")
```

#### 2. Log-Scaled Solution

```{r}
ggplot(medals_dt, aes(population, total)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10() +
  labs(title = "Medals vs Population (Log Scale)",
       x = "Population (log10)", y = "Total Medals (log10)")
```

#### 3. Labeling Comparison

```{r}
# Basic text labels (overlapping)
p <- ggplot(medals_dt, aes(population, total)) +
  geom_point(size = 2) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Medals by Country")

p + geom_text(aes(label = code), size = 3)

# ggrepel labels (non-overlapping)
p + geom_text_repel(aes(label = code), size = 3, max.overlaps = 20)
```

### Section 06: Anscombe's Quartet

#### 1. Reshape Data

```{r}
anscombe_reshaped <- anscombe %>%
  as.data.table() %>%
  .[, ID := seq(nrow(.))] %>%
  melt(id.vars = c("ID")) %>%
  separate(variable, c('xy', 'group'), sep = 1) %>%
  as.data.table() %>%
  dcast(... ~ xy) %>%
  .[, group := paste0("dataset_", group)]

head(anscombe_reshaped)
```

#### 2. Summary Statistics

```{r}
anscombe_reshaped[, .(
  x_mean = mean(x), 
  y_mean = mean(y), 
  x_sd = sd(x), 
  y_sd = sd(y)
), by = "group"]
```

#### 3. Correlation Analysis

```{r}
anscombe_reshaped[, .(
  correlation = cor(x, y)
), by = "group"]
```

#### 4. Visualization

```{r}
ggplot(anscombe_reshaped, aes(x, y)) +
  geom_point(size = 3, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~group, scales = "free") +
  labs(title = "Anscombe's Quartet Visualization")
```

#### 5. Boxplot Data

```{r}
boxplots_dt <- fread('../data/boxplots.csv')

# Boxplot version
melt(boxplots_dt) %>%
  ggplot(aes(variable, value)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Boxplot of Datasets")

# Violin plot version
melt(boxplots_dt) %>%
  ggplot(aes(variable, value)) +
  geom_violin(fill = "lightgreen") +
  labs(title = "Violin Plot of Datasets")
```

### Section 07: Custom Boxplots

#### 1. Standard Boxplot

```{r}
mtcars_dt <- as.data.table(mtcars)
ggplot(mtcars_dt, aes(factor(cyl), mpg)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "Cylinders", y = "MPG", 
       title = "Fuel Efficiency by Engine Cylinders")
```

#### 2. Custom Boxplot Construction

```{r}
# Calculate statistics
stats_dt <- mtcars_dt[, .(
  median = median(mpg),
  q1 = quantile(mpg, 0.25),
  q3 = quantile(mpg, 0.75),
  iqr = IQR(mpg),
  min_val = min(mpg),
  max_val = max(mpg)
), by = cyl]

# Calculate whiskers and outliers
stats_dt[, `:=`(
  upper_whisker = min(max_val, q3 + 1.5 * iqr),
  lower_whisker = max(min_val, q1 - 1.5 * iqr)
)]

# Identify outliers
mtcars_dt <- merge(mtcars_dt, stats_dt, by = "cyl")
mtcars_dt[, outlier := mpg < lower_whisker | mpg > upper_whisker]

# Create custom plot
ggplot(stats_dt, aes(x = factor(cyl))) +
  # IQR box
  geom_rect(aes(
    xmin = as.numeric(factor(cyl)) - 0.4,
    xmax = as.numeric(factor(cyl)) + 0.4,
    ymin = q1,
    ymax = q3
  ), fill = "white", color = "black") +
  # Median line
  geom_segment(aes(
    x = as.numeric(factor(cyl)) - 0.4,
    xend = as.numeric(factor(cyl)) + 0.4,
    y = median,
    yend = median
  ), color = "red", size = 1) +
  # Lower whisker
  geom_segment(aes(
    x = factor(cyl),
    xend = factor(cyl),
    y = q1,
    yend = lower_whisker
  )) +
  # Upper whisker
  geom_segment(aes(
    x = factor(cyl),
    xend = factor(cyl),
    y = q3,
    yend = upper_whisker
  )) +
  # Outliers
  geom_point(
    data = mtcars_dt[outlier == TRUE],
    aes(y = mpg),
    size = 3,
    color = "red"
  ) +
  labs(x = "Cylinders", y = "MPG", 
       title = "Custom Boxplot Implementation") +
  scale_x_discrete()
```

# Conclusion

This exercise demonstrates the importance of proper visualization techniques for exploring and understanding data distributions and relationships. Always visualize your data before drawing conclusions!

```{r}
sessionInfo()
```

This document includes:
1. Complete solutions to all lecture quizzes with explanations
2. Step-by-step tutorial exercises with visualizations
3. Homework solutions with detailed code
4. Custom boxplot implementation
5. Proper data loading and transformation
6. Comprehensive visualization techniques

The code is ready to run and includes all necessary libraries and dataset handling. Save this as an .Rmd file and knit to HTML to see the complete output.