---
title: "08_statistical_testing_3"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  cache = TRUE
)
library(ggplot2)
library(data.table)
library(magrittr)
library(tidyr)
library(dplyr)
library(datasets)
```

# Tutorial Solutions

## Section 01 - Choosing the Right Test

```{r}
# (a) Spearman correlation (quantitative, no distributional assumptions)
# (b) Paired t-test (quantitative, normal distribution)
# (c) Binomial test (H0: p=0.99, alternative="less")
# (d) Fisher's exact test (binary-binary association, alternative="greater")
```

## Section 02 - Marker-Growth Association

### Load Data

```{r}
gene <- fread("../data/eqtl/gene.txt")
genotype <- fread("../data/eqtl/genotype.txt") %>% 
  melt(id.vars = 'strain', variable.name = 'marker', value.name = 'genotype')
growth <- fread("../data/eqtl/growth.txt") %>% 
  melt(id.vars = "strain", variable.name = 'media', value.name = 'growth_rate')
marker <- fread("../data/eqtl/marker.txt")
```

### Helper Functions

```{r}
getMaltoseDt <- function(mrk){
  merge(growth[media == "YPMalt"], 
        genotype[marker == mrk, .(strain, genotype)], 
        by = 'strain')
}

plot_growth_one_mk <- function(mk){
  ggplot(getMaltoseDt(mk), aes(genotype, growth_rate)) +
    geom_boxplot() +
    labs(title = mk) + 
    theme_bw(base_size = 16) +
    theme(plot.title = element_text(hjust = 0.5))
}
```

### Marker 5211 Analysis

```{r}
#| fig-height: 4
plot_growth_one_mk("mrk_5211")

# Statistical tests
m_dt <- getMaltoseDt('mrk_5211')
t_test <- t.test(growth_rate ~ genotype, m_dt)
wilcox_test <- wilcox.test(growth_rate ~ genotype, m_dt)

cat("t-test p-value:", format.pval(t_test$p.value), "\n")
cat("Wilcoxon p-value:", format.pval(wilcox_test$p.value))
```

### Test Function

```{r}
test_growth <- function(mk, test){
  m_dt <- getMaltoseDt(mk)
  if(test == 'wilcoxon') {
    wilcox.test(growth_rate ~ genotype, m_dt)$p.value
  } else {
    t.test(growth_rate ~ genotype, m_dt)$p.value
  }
}

# Marker 1653 (non-significant)
test_growth('mrk_1653', 'wilcoxon')  # ~0.50
test_growth('mrk_1653', 't')         # ~0.50

# Marker 5091 (significant)
test_growth('mrk_5091', 'wilcoxon')  # ~0.0008
test_growth('mrk_5091', 't')         # ~0.0006
```

## Section 03 - Correlation Analysis

### Full Iris Dataset

```{r}
#| fig-height: 5
data(iris)
cor_test_pearson <- cor.test(iris$Sepal.Length, iris$Sepal.Width, method = "pearson")

ggplot(iris, aes(Sepal.Length, Sepal.Width)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(title = paste("Pearson r =", round(cor_test_pearson$estimate, 3),
                     "p =", format.pval(cor_test_pearson$p.value)))
```

### By Species

```{r}
#| fig-height: 6
iris_dt <- as.data.table(iris)
corr_dt <- iris_dt[, {
  test <- cor.test(Sepal.Length, Sepal.Width, method = "pearson")
  list(estimate = test$estimate, p.value = test$p.value)
}, by = Species]

ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point() +
  geom_smooth(method = lm) +
  geom_text(data = corr_dt, aes(x = 6.5, y = 4.2, 
                                label = paste("r =", round(estimate, 2))), 
            color = "black") +
  facet_wrap(~Species) +
  theme_bw(base_size = 12)
```

## Section 04 - Normality Assumption Pitfalls

```{r}
#| fig-height: 10
dt <- fread("../data/stats-pitfalls.csv")

# Visualization
p1 <- ggplot(melt(dt), aes(variable, value)) + 
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", color = "red", size = 3) +
  labs(title = "Boxplots with Mean (red dot)")

p2 <- ggplot(melt(dt), aes(value)) + 
  geom_histogram(bins = 20) +
  facet_wrap(~variable, scales = "free") +
  geom_vline(aes(xintercept = mean(value)), color = "red") +
  labs(title = "Histograms with Mean (red line)")

gridExtra::grid.arrange(p1, p2, ncol = 1)

# Statistical tests
cat("Wilcoxon p-value:", wilcox.test(dt$group1, dt$group2)$p.value, "\n")
cat("t-test p-value:", t.test(dt$group1, dt$group2)$p.value)
```

## Section 05 - Correlation Test

```{r}
ex_dt <- fread("../data/exam_correlation.csv")

# Correlations
pearson_cor <- cor.test(ex_dt$attendance, ex_dt$achieved_points, method = "pearson")
spearman_cor <- cor.test(ex_dt$attendance, ex_dt$achieved_points, method = "spearman")

# Visualization
ggplot(ex_dt, aes(attendance, achieved_points)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(title = paste("Pearson r =", round(pearson_cor$estimate, 3), 
                    "| Spearman Ï =", round(spearman_cor$estimate, 3)),
       subtitle = "Spearman preferred due to outliers (0 scores)")
```

## Section 06 - Association Tests

```{r}
# Mtcars analysis
cars <- as.data.table(mtcars)
cars[, cyl_4 := cyl > 4]
cars[, gear_3 := gear > 3]
cont_tbl <- table(cars$cyl_4, cars$gear_3)

# Two-sided Fisher test
signif(fisher.test(cont_tbl)$p.value, 2)  # 0.0028

# Interpretation: Significant association between high cylinders and gears
```

## Genetic Linkage Analysis

```{r}
# Marker association function
marker_test <- function(marker1, marker2){
  mks_geno <- genotype[marker %in% c(marker1, marker2)] %>%
    dcast(strain ~ marker, value.var = "genotype")
  fisher.test(table(mks_geno[[2]], mks_geno[[3]]))$p.value
}

# Test all markers against mrk_1
other_markers <- marker[id != "mrk_1"]
other_markers[, pval := sapply(id, marker_test, "mrk_1")]
other_markers[, on_chr_1 := chrom == "chr01"]

# Visualization
ggplot(other_markers, aes(start, -log10(pval))) +
  geom_point() +
  facet_wrap(~chrom, scales = "free_x", nrow = 2) +
  labs(x = "Genomic Position", y = "-log10(P-value)")

# P-value distribution by chromosome
ggplot(other_markers, aes(pval)) +
  geom_histogram(bins = 20) +
  facet_wrap(~on_chr_1, labeller = label_both) +
  labs(title = "P-value Distribution by Chromosome Group")

cat("Chr1 significant fraction:",
    mean(other_markers[on_chr_1 == TRUE]$pval < 0.05), "\n")
cat("Other chromosomes significant fraction:",
    mean(other_markers[on_chr_1 == FALSE]$pval < 0.05))
```

### Key Observations

- Markers on chr01 show strong distance-dependent association with mrk_1
- Non-chr01 markers show baseline ~5% significant rate (false positives)
- Demonstrates genetic linkage on same chromosome